Plastic in the Urban Freshwater Environment

### 2021-11-13 products/manuscript/Manuscript.Rmd
results

### Pollution Sources
Microplastics are classified by source as either primary or secondary. Primary microplastics are those which are originally manufactured as microplastics, such as microbeads; primary microplastics are often found in personal care products, industrial scrubbers used for cleaning, plastic powder used for molding, and virgin resin used in plastic production (EFSA, 2016). Secondary microplastics begin as larger plastics and degrade into smaller pieces as the result of exposure to environmental stressors (e.g. sunlight, wind, water) (Murphy et al., 2016). For example, secondary microplastics occur when ultraviolet light exposure catalyzes photo-oxidation in plastic, leading to brittle pieces that are even more likely to break down (Wright & Kelly, 2017). A secondary microplastic source that is less obvious than the fragmentation of plastic debris is synthetic clothing. Clothes made of fibers such as nylon, polyester, rayon, or spandex shed microfibers during washing cycles; these fibers are later deposited into the environment through wastewater treatment plant effluent (Murphy et al., 2016). Most experts in the field agree that secondary microplastics are the most significant contributor to microplastic pollution, though a broad range of microplastic sources are recognized (Tibbetts et al., 2018). Researchers recognize wastewater treatment plant (WWTP) effluent as a major vector for both primary and secondary microplastics, and therefore WWTP effluent has been one area of focus in microplastics literature (Murphy et al., 2016; McCormick et al., 2016; Li et al., 2018, Tibbetts et al., 2018, Harrison et al., 2018). A team of scientists in Scotland examined WWTPs as a source of microplastics in the aquatic environment by sampling for microplastics at different stages of wastewater treatment; the team calculated a 98% reduction of microplastics from influent to effluent (Murphy et al., 2016). While an efficient removal rate, Murphy and collaborators estimate that 65 million microplastics are still released from the plant daily due to the large volume of effluent. Numerous other studies have evaluated WWTP effluent in relation to microplastic concentration as well (McCormick et al., 2016; Li et al., 2018, Tibbetts et al., 2018, Harrison et al., 2018). 
Other predictors of microplastics particle concentration that researchers consider include land use (i.e. natural, agricultural, industrial, developed, forest & woodland, etc.) and population (total size, density, dwellings) (Tibbetts et al., 2018; Barrows et al, 2018; Townsend et al., 2019; Yonkos et al, 2014). Compared to the ocean, freshwater is the dominant source of microplastic due to land-based industrial and domestic pollution (Luo et al., 2019). Rivers transport both point-source and diffuse pollution to the ocean, contributing an estimated 1.15 to 2.41 million tons of plastic waste annually (Lebreton et al, 2017). Factors including anthropogenic impact, pollution sources, and hydrodynamics affect the rate of microplastic accumulation and transport (Luo et al., 2019; CITE). Researchers continue to focus on identifying the variables influencing microplastic concentration in present day research (CITE).

### Ecological Consequences
Despite their small size, microplastics have significant and quantifiable impact on aquatic ecosystems (CITE). Since microplastics permeate both marine and freshwater environments, aquatic animals easily mistake the particles for food and ingest them, allowing the pollutant to accumulate in their bodies; studies have shown that organisms ingest microplastics either by accident or by mistaking them for food (Wright & Kelly, 2017; Carbery, O’Connor, & Palanisami, 2018, Tanaka & Takada, 2016). Microplastics have been identified in the digestive tracts of fish and birds of prey, in mussels and other bivalves, and in natural populations of a crustacean species (Tanaka & Takada, 2016; Carlin et al., 2020; van Cauwenberghe & Janssen, 2014; Devriese et al., 2015; Murray and Cowie, 2011). Microplastics are dangerous to organisms due to the physical accumulation of the particles themselves, and because they act as a transport mechanism for toxic chemicals in three main ways: (1) the plastic itself is toxic, (2) the plastic adsorbs and later releases environmental pollutants, and (3) the plastic leaches chemical additives (Mato et al, 2001; Bouwmeester et al., 2015). The hydrophobic surface and high surface area to volume ratio of microplastics causes the adsorption and concentration of organic pollutants such as polychlorinated bisphenols (PCBs), dichlorodiphenyldichloroethylene (DDE), and nonylphenol (NP) (Wright & Kelly, 2017; Carbery, O’Connor, & Palanisami, 2018). Evidence of trophic transfer of microplastics through the food chain suggests that microplastics and the toxins that accompany them pose a threat to aquatic life further up the food chain, and to human health through consumption of seafood (Barboza et al 2018, Carbery et al, 2018).

@@

### Present Research
	This preliminary study sought to identify the presence of microplastics in the Upper Oconee Watershed, estimate the levels of microplastic pollution present in the freshwater systems of Athens, GA, observe temporal changes and trends, and explore the relationship between microplastic pollution and hypothesized predictors including population level, land use, and wastewater treatment plant proximity (CITE – who hypothesizes this?). We hypothesized that microplastics would be present in the Upper Oconee Watershed, that levels would be greater in more highly-trafficked areas with greater anthropogenic impact, and that microplastic pollution would correlate to other indicators of water quality (CITE).

@@

Samples were processed via vacuum filtration, which was performed in 150 mL volumes in duplicate for each site. Samples were pumped through gridded 0.45 micron membranes (Whatman) and immediately covered and stored covered in plastic petri dishes at room temperature for up to 8 weeks. 
Filters were visually inspected using a dissecting microscope with magnification ranging from [INSERT]. Microplastic particles were identified based on the following characteristic features: unnatural homogenous color, unnatural shape, homogenous texture, homogenous width (no tapered end) (CITE). Microplastic concentration (No/L) was calculated by dividing the total microplastic particle count by the sample volume. Duplicates (samples A and B for each site) were averaged to obtain a singular concentration value for each sample site.

@@

@@

>>>>

### Pollution Sources
Microplastics are classified by source as either primary or secondary. Primary microplastics are those which are originally manufactured as microplastics, such as microbeads; primary microplastics are often found in personal care products, industrial scrubbers used for cleaning, plastic powder used for molding, and virgin resin used in plastic production [@EFSA2016]. Secondary microplastics begin as larger plastics and degrade into smaller pieces as the result of exposure to environmental stressors (e.g. sunlight, wind, water) [@Murphy2016]. For example, secondary microplastics occur when ultraviolet light exposure catalyzes photo-oxidation in plastic, leading to brittle pieces that are even more likely to break down [@Wright2017]. A secondary microplastic source that is less obvious than the fragmentation of plastic debris is synthetic clothing. Clothes made of fibers such as nylon, polyester, rayon, or spandex shed microfibers during washing cycles; these fibers are later deposited into the environment through wastewater treatment plant effluent [@Murphy2016]. Most experts in the field agree that secondary microplastics are the most significant contributor to microplastic pollution, though a broad range of microplastic sources are recognized [@Tibbetts2018]. Researchers recognize wastewater treatment plant (WWTP) effluent as a major vector for both primary and secondary microplastics, and therefore WWTP effluent has been one area of focus in microplastics literature [@Murphy2016;@McCormick2014a;@Li2018;@Tibbetts2018;@Harrison2018]. A team of scientists in Scotland examined WWTPs as a source of microplastics in the aquatic environment by sampling for microplastics at different stages of wastewater treatment; the team calculated a 98% reduction of microplastics from influent to effluent [@Murphy2016]. While an efficient removal rate, Murphy and collaborators estimate that 65 million microplastics are still released from the plant daily due to the large volume of effluent. Numerous other studies have evaluated WWTP effluent in relation to microplastic concentration as well [@McCormick2014a;@Li2018,@Tibbetts2018,@Harrison2018]. 
Other predictors of microplastics particle concentration that researchers consider include land use (i.e. natural, agricultural, industrial, developed, forest & woodland, etc.) and population (total size, density, dwellings) [@Tibbetts2018; @Barrows2018;@Townsend2019; @Yonkos2014]. Compared to the ocean, freshwater is the dominant source of microplastic due to land-based industrial and domestic pollution [@Luo2019]. Rivers transport both point-source and diffuse pollution to the ocean, contributing an estimated 1.15 to 2.41 million tons of plastic waste annually [@Lebreton2017]. Factors including anthropogenic impact, pollution sources, and hydrodynamics affect the rate of microplastic accumulation and transport [@Luo2019; CITE]. Researchers continue to focus on identifying the variables influencing microplastic concentration in present day research (CITE).

### Ecological Consequences
Despite their small size, microplastics have significant and quantifiable impact on aquatic ecosystems (CITE). Since microplastics permeate both marine and freshwater environments, aquatic animals easily mistake the particles for food and ingest them, allowing the pollutant to accumulate in their bodies; studies have shown that organisms ingest microplastics either by accident or by mistaking them for food [@Wright2017;@Carbery2018;@Tanaka2016]. Microplastics have been identified in the digestive tracts of fish and birds of prey, in mussels and other bivalves, and in natural populations of a crustacean species [@Tanaka2016;@Carlin2020;@VanCauwenberghe2014;@Devriese2015;@Murray2011]. Microplastics are dangerous to organisms due to the physical accumulation of the particles themselves, and because they act as a transport mechanism for toxic chemicals in three main ways: (1) the plastic itself is toxic, (2) the plastic adsorbs and later releases environmental pollutants, and (3) the plastic leaches chemical additives [@Mato2001;@Bouwmeester2015]. The hydrophobic surface and high surface area to volume ratio of microplastics causes the adsorption and concentration of organic pollutants such as polychlorinated bisphenols (PCBs), dichlorodiphenyldichloroethylene (DDE), and nonylphenol (NP) [@Wright2017;@Carbery2018]. Evidence of trophic transfer of microplastics through the food chain suggests that microplastics and the toxins that accompany them pose a threat to aquatic life further up the food chain, and to human health through consumption of seafood [@Barboza2018;@Carbery2018].

@@

### Present Research
This preliminary study sought to identify the presence of microplastics in the Upper Oconee Watershed, estimate the levels of microplastic pollution present in the freshwater systems of Athens, GA, observe temporal changes and trends, and explore the relationship between microplastic pollution and hypothesized predictors including population level, land use, and wastewater treatment plant proximity (@Murphy2016). We hypothesized that microplastics would be present in the Upper Oconee Watershed, that levels would be greater in more highly-trafficked areas with greater anthropogenic impact, and that microplastic pollution would correlate to other indicators of water quality [@McCormick2014a].

@@

Samples were processed via vacuum filtration, which was performed in 150 mL volumes in duplicate for each site. Samples were pumped through gridded 0.45 micron membranes (Whatman) and immediately covered and stored covered in plastic petri dishes at room temperature for up to 8 weeks. 
Filters were visually inspected using a dissecting microscope with magnification ranging from [INSERT]. Microplastic particles were identified based on the following characteristic features: unnatural homogenous color, unnatural shape, homogenous texture, homogenous width (no tapered end)[@Hidalgo-Ruz2012]. Microplastic concentration (No/L) was calculated by dividing the total microplastic particle count by the sample volume. Duplicates (samples A and B for each site) were averaged to obtain a singular concentration value for each sample site.

@@

### Mapping
I created some maps for concentrations at each sample sites throughout the seasons (saved as html files "map_season.html") for now, but I am unsure of how best to display these maps in the manuscript word document output. 
```{r}
#mapview::knit_print.mapview(here("results", "map_spring.html"))
```

@@

Based on the results of the three different models, the LASSO model is the best option for this dataset, though the minimal difference in RMSE when compared to the null model suggests that even though LASSO is the better model method compared to others, it still does not produce a great model for predicting microplastic concentration.
